Identifying Inter-Genomic Repeats Using Fast, Approximate Methods for Betweenness Centrality

### 2015-05-18 manuscript/manuscript.tex
Revised intro.

Metagenomic data are considerably more complex. The assembly of metagenomic data is complicated due to several reasons. Few of them are: (i) widely different levels of representation for different organisms in a community; (ii) genomic variation between closely related organisms; (iii) conserved genomic regions shared by distantly related organims; and (iv) repetitive sequences within individual genomes. Genomic repeats are the major challenge when assembling isolated genomes and their effect is prominently seen in metagenomic datasets. It has been shown that the number of reconstructions grow exponentially with the number of repeats\cite{kingsford}. Therefore it is computationally infeasible to find one correct solution for assembly.  To tackle this problem, several assemblers have used the notion of network centrality to find repeats. For example, Bambus2\cite{bambus} uses betweenness centrality to find central nodes in graphs. The drawback of their approach is that they calculate exact betweenness centrality by computing all pair shortest paths. This approach is very slow when applied to large assembly graphs. To overcome this drawback, we explore how can we apply approximate betweenness centrality to assembly graphs and find inter-genomic repeats accurately. The reason for using approximation algorithms is that they scale well as the size of graphs increase. We also investigate how parallelized implementation of these algorithms provide speedups in detecting repeats.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{es_mix_200kb_k21.png}
\caption{Assembly graph of a simulated community consisting of 200 Kbp subsets of \textit{Escherichia coli} str. K-12 MG1655 and \textit{Staphylococcus aureus}.  Nodes are colored and sized based on their relative betweenness centrality. The highlighted nodes are inter-genomic repeats whose deletion would separate the graph.}
\label{fig:sampled_nodes}
\end{figure*}

@@

Another cutoff criteria we test is based on \textit{interquartile range(IQR)}. Let $Q_{1}$,$Q_{2}$, and $Q_{3}$ denote lower, middle and upper quartile respectiely. IQR can be defined as the difference between the upper and lower quartile. So, $IQR = Q_{3} - Q_{1}$. For the values of centrality, we calculate upper quartile, middle quartile, lower quartile and IQR. For filtering, we mark all the nodes with centrality more than $Q_{3} + 1.5*IQR$ as repeats. The advantage of this method is that it does not make any assumptions about the distribution of underlying data.

@@

Decreasing the betweenness centrality cutoff requirement results in a trade-off between the different statistical measures.
Instead of using the default cutoff of $3\sigma$ suggested by Bambus2, we mark all the nodes with centrality more than $Q_{3} + 1.5*IQR$ as an inter-genomic repeats.
In the 10-genome assembly graph, the sensitivity increases from 5.8\% to 34.4\% when using the \textsc{iqr} cutoff method.
Conversely, the specificity and and precision drops from 99.3\% to 90.9\% and 25\% to 12.5\%, respectively.

>>>>

The assembly of metagenomics data, the sequences of DNA recovered from an environment without culturing, is considerably more complex than that of individual genomes.
%Challenges include: (i) repetitive sequences within individual genomes; (ii) varying abundances of different organisms in the community; (iii) variation between closely related organisms; and (iv) conserved genomic segments shared by distantly related organisms.
Repetitive sequences pose a major challenge when assembling isolated genomes.
It has been shown that the number of reconstructions grow exponentially with the number of repeats making it computationally infeasible to find one correct solution for assembly\cite{kingsford}.
The effect of repeats on assembly is exacerbated in metagenomic data sets due to the presence of conserved regions of the genome linking together distantly related organisms.
To tackle the problem of identifying these inter-genomic repeats, several assemblers have used the notion of network centrality to find repeats within the assembly graph.
For example, Bambus2\cite{bambus} uses betweenness centrality to find \emph{central} nodes in the graph.
The drawback of their approach is the expensive calculation of betweenness centrality by computing all-pair shortest paths.
This computation is impractical when applied to the large assembly graphs typically encountered.
To overcome this drawback, we explore the application of approximate and parallel betweenness centrality algorithms on metagenomic assembly graphs and evaluate their effectiveness in identifying inter-genomic repeats.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{es_mix_200kb_k21.png}
\caption{Assembly graph of a simulated community consisting of 200 Kbp subsets of \textit{Escherichia coli} str. K-12 MG1655 and \textit{Staphylococcus aureus}.  Nodes are colored and sized based on their relative betweenness centrality. The highlighted nodes are inter-genomic repeats whose deletion would separate the graph. Note that the betweenness centrality measure correctly identifies these nodes.}
\label{fig:sampled_nodes}
\end{figure*}

@@

Another cutoff criteria we test is based on \textit{interquartile range(IQR)}. Let $Q_{1}$,$Q_{2}$, and $Q_{3}$ denote the lower, middle and upper quartiles respectively. IQR can be defined as the difference between the upper and lower quartile. So, $IQR = Q_{3} - Q_{1}$. For the values of centrality, we calculate upper quartile, middle quartile, lower quartile and IQR. For filtering, we mark all the nodes with centrality more than $Q_{3} + 1.5*IQR$ as repeats. The advantage of this method is that it does not make any assumptions about the distribution of underlying data.

@@

Decreasing the betweenness centrality cutoff requirement results in a trade-off between the different statistical measures.
Instead of using the default cutoff of $3\sigma$ suggested by Bambus2, we mark all the nodes with centrality more than $Q_{3} + 1.5*IQR$ as an inter-genomic repeats.
In the 10-genome assembly graph, the sensitivity increases from 5.8\% to 34.4\% when using the IQR cutoff method.
Conversely, the specificity and and precision drops from 99.3\% to 90.9\% and 25\% to 12.5\%, respectively.
